Canonical rings of rational divisors on projective space and Hirzebruch surfaces

### 2015-07-15 surfaces-write-up/surface-write-up.tex

\noindent
Let $e_1, \ldots, e_t \in \Sigma,$ with $e_i = (\pdeg_i, c_1^i,
\ldots, c_n^i)$ be a set of {\bf extremal rays} of $\Sigma$,
meaning that the $\Sigma$ is contained in the $R_{\geq 0}$ span of
$e_1, \ldots, e_t$.  Then, as a semigroup, $\Sigma$ is generated by
elements whose first coordinate is less than $\sum_{i = 1}^{n} \pdeg_i$
. Furthermore, every element $\sigma \in \Sigma$ can be written in
a canonical form $\lambda + \sum_{i=1}^{t} w_i e_i,$ with the first
coordinate of $\lambda$ less than $\sum_{i=1}^{t}\pdeg_i$ and $w_1,
\ldots, w_t \in \bz$.
\end{prop}

@@

\begin{lem}
\label{lem:bound-ker-chi}
Retaining the notation of Lemma ~\ref{lem:composite-map}, if $\Sigma$ has extremal rays $e_1,\ldots, e_r$ in degrees $d_1, \ldots, d_r$ then $\ker \chi$ has a system of generators in degrees less than $2(\sum_{i=1}^{r}d_r-1)$.
\end{lem}
\begin{proof}
Since $e_1, \ldots, e_r$ are extremal rays, by Proposition ~\ref{prop:cone-generation}, every element $\sigma \in \Sigma$ can be written in a canonical form $\lambda + \sum_{i=1}^{r}e_i$. Now, letting $\lambda_0 := 0,\lambda_1, \ldots, \lambda_m$ be all elements of $\Sigma$ so that we can write $\lambda_i = \sum_{i=1}^{r}\bida_i e_i$ with $0 \leq \bida_i < 1$ for all $i$. Then, for any $1 \leq i \leq j \leq m,$ we can write $\lambda_i + \lambda_j$ in the above canonical form, yielding a relation in degree at most $\deg \lambda_i + \deg \lambda_j \leq 2 \cdot \left( \sum_{i=1}^{r}d_r -1 \right).$ Furthermore, these relations generate all relations, as one can apply a sequence of these relations to put any $\sigma \in \Sigma$ into canonical form.
\end{proof}

@@

\begin{rem}
\label{rem:ghost-motivation}
Example ~\ref{eg:hyperplane} shows that the naive generalization of ~\ref[Theorem 8]{dorney:canonical} of generation in degree at most $\sum_{i=1}^{n}\ell_i$ cannot possibly hold. The reason for this is that the divisors may ``be contained in a hyperplane,'' i.e., be expressable as functions in $m$ of the $m+1$ variables on $\bp^m$. This problem can easily be circumvented by adding in ``ghost points.'' That is, we may add divisors of the form $0 \cdot H_i$ to $D$, and reorder so that if $D = \sum_{i=1}^{n}\alpha_i V(f_i)$, then $f_0, \ldots, f_n$
\end{rem}

@@

Let $D = \sum_{i=1}^n \alpha_i C_i$ for curves $C_i$ on $\mathbb{P}^m$.  Let $f_i$ be homogeneous polynomials in $x_0, \ldots, x_m$ such that $C_i = v(f_i)$ for all $i$; suppose $\deg(f_i) = a_i$. 
\begin{lem} \label{P-m-generators}
$R_D$ is generated in degree at most $\sum_{i=1}^n \ell_i a_i$.
% and relations in degree at most  .\todo{insert numbers}
\end{lem}
\begin{proof}
Let 
\begin{align}\label{eqn:Sigma-def}
	\Sigma = \{(d, c_1, \ldots, c_n) \in \bz^{n+1} : c_i \geq - d \alpha_i,1 \leq i \leq n \text{ and } \sum_{i=1}^{n}k_i^j = 0, 1 \leq j \leq t\}.
\end{align}
We note that $\Sigma$ has extremal rays given by the lattice points $e_i$ corresponding to
\begin{equation}\label{eqn:epilon-def-P-m}
	\epsilon_i = \prod_{j\ne i} (f_j)^{-\alpha_j \ell_i a_i} {f_i}^{\ell_i \sum_{j\ne i} \alpha_j a_j}
\end{equation}
for each $i\in \{1, \ldots n\}$.  Therefore, applying proposition \ref{prop:cone-generation}, we see $R_D$ is generated in degree less than
\[
	\sum_{i=1}^n \ell_i a_i.
\]
\end{proof}

Let $x_1, \ldots x_r$ be the generators given by lemma  \ref{P-m-generators}, and $\phi$ be the surjection $k[x_1, \ldots x_r] \to R_D$.  We can factor $\phi$ through 
\[
	\Sigma =  \langle u^d z_1^{c_1} \cdots z_n^{c_n} : c_i \geq -d \alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{n} \bidb_i c_i. \rangle 
\]
into
\[
\begin{tikzcd}
\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
x_i \ar r & u^{d_i}z_1^{c_{i1}} \cdots z_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]

Applying Lemma \ref{lem:bound-ker-chi}, we can bound the degree of generation of $\ker(\chi)$ less than
\[
	\sum_{i=1}^n \ell_i a_i.
\]
Finally, we calculate the degree of generation of $\psi$:

\begin{lem}\label{P-m-relations-psi}
$\ker(\psi)$ is generated in degree at most 
\[
	\delta_i + \sum_{i=1}^n \ell_i a_i.
\]
\end{lem}
\begin{proof}
We first claim that $\psi$ has a relation of the form
\begin{equation}\label{eqn:relations-psi}
	(z_i - \beta_j)\prod_{j=1}^n {z_j}^{c_{j}}
\end{equation}

\noindent
for some $d\in \mathbb{N}$, $c_i \ge -\alpha_i d$, $\bida_j + \sum_{i = 1}
^n \bida_i c_i = 0$, $\bidb_j + \sum_{i=1}^n \bidb_i c_i = 0$, and $\beta_j \in
k[z_1, \ldots, y_3]$ satisfies $\psi(\beta_j)\in \mathscr{O}(\bida)$.

To find such a $\beta$, we simply write $\epsilon_j$ as a 
polynomial in $\epsilon_1, \ldots \epsilon _3$. Furthermore, these
generate all relations, since they allow us to reduce any $\prod_{i =
1}^n \epsilon_i^{r_i}$ to a canonical form. \todo{add reference to
earlier description of canonical form.}

Now we bound the degree of generation of these relations for a fixed $z_i$.  We associate $(z_i - \beta_j)\prod_{j=1}^n z_j^{c_j}$ that has degree $d$ with the lattice point $(d, c_1, \ldots, c_n)$.  Let $\Sigma_i \subseteq \mathbb{R}^{n+1}$ be the set of points of the form $(d, r_1, \ldots, r_n)$ satisfying $r_j \ge -d \alpha_j$ for all $j$ and $\sum_{j=1}^n r_j a_j = -a_i$.  Then we see $\Sigma_i$ is a translation of $\Sigma$ (see Equation \ref{eqn:Sigma-def}) by $\delta_i = (\frac{a_i}{\deg(D)}, -\frac{\alpha_1 a_1}{\deg(D)}, \ldots, - \frac{\alpha_n a_n}{\deg(D)})$.

Therefore, we can write any element of $\Sigma'$ as
\[
	\delta_i + \sum_{j=1}^n r_j e_j
\]
for $r_j \in \mathbb{R}$ for each $j$.

Whenever some $r_j \ge 1$, we can generate the generation at hand by one of lower degree by factoring out the section $\epsilon_j$ corresponding to the lattice point $e_j$ (see Equation \ref{eqn:epilon-def-P-m}).  Therefore, we can bound the degree of generation of $\ker(\psi)$ less than
\[
	\delta_i + \sum_{i=1}^n \ell_i a_i.
\]
\end{proof}

Therefore, since $\ker(\chi)$ and $\ker(\psi)$ are generated in degrees less than $\sum_{i=1}^n \ell_i a_i$ and $\delta_i + \sum_{i=1}^n \ell_i a_i$ respectively, Lemma \ref{lem:composite-map} implies that $\ker(\phi)$ is generated in degrees less than
\[
	\max \left(\sum_{i=1}^n \ell_i a_i, \delta_i + \sum_{i=1}^n \ell_i a_i \right).
\]

@@

\begin{thm}\label{P-m-generators/relations}
Let $D = \sum_{i=1}^n \alpha_i C_i$ for curves $C_i$ on $\mathbb{P}^m$.  Let $f_i$ be homogeneous polynomials in $x_0, \ldots, x_m$ such that $C_i = v(f_i)$ for all $i$; suppose $\deg(f_i) = a_i$. 
Then $R_D$ is generated in degree at most 
\[
	\sum_{i=1}^n \ell_i a_i
\]
with relations generated in degree at most
\[
	\max \left(\sum_{i=1}^n \ell_i a_i, \delta_i + \sum_{i=1}^n \ell_i a_i \right).
\]
\end{thm}

\section{Canonical Rings on Hirzebruch surfaces}
Let $D=\sum_{i=1}^m \alpha_i C_i \in \mathbb{Q} \otimes \di(\hirz_
k)$ for some curve $C_i$ of $\hirz_k$, such
that $C_1, \ldots, C_4$ are independent curves with bi-degrees $(1,0)
, (1,0), (0,1)$, and $(0,1)$ respectively. Let $C_i = V(f_i)$ where
$f_i \in \mathscr{O}(\bida_i, \bidb_i)$.  Suppose $f_1$ and $f_2$ are independent linear
polynomials in $x_0$ and $x_1$ and $f_3$ and $f_4$ are independent
linear polynomials in $y_0$ and $y_1$.

\begin{defn}
Define 
\begin{equation}\label{eqn:define-T=(D)}
	\Te(D) = \left\{i \in \{1, \ldots, m\}: \bida_i \sum_{k=1}^m \bidb_k 
\alpha_k = \bidb_i \sum_{k=1}^m \bida_k \alpha_k \right\},
\end{equation}

\begin{equation}\label{eqn:define-T+(D)}
	\Tp(D) = \left\{ i \in \{1, \ldots, m\}:  \vphantom{\sum_{k=1}^m} 
\bida_i \sum_{k=1}^m \alpha_k \bidb_k > \bidb_i \sum_{k=1}^m \alpha_k \bida_k 
\right\},
\end{equation}

\noindent
and
\begin{equation}\label{eqn:define-T-(D)}
	\Tm(D) = \left\{ j \in \{1, \ldots, m\}: \bida_j \sum_{k=1}^m \alpha_
k \bidb_k < \bidb_j \sum_{k=1}^m \alpha_k \bida_k s \right\}.
\end{equation}
\end{defn}

>>>>

\noindent
Suppose $e_1, \ldots, e_t \in \Sigma$ with $e_i = (\pdeg_i, c_1^i,
\ldots, c_n^i)$ are a set of {\bf extremal rays} of $\Sigma$,
meaning that the $\Sigma$ is contained in the $R_{\geq 0}$ span of
$e_1, \ldots, e_t$.  Then, as a semigroup, $\Sigma$ is generated by
elements whose first coordinate is less than $\sum_{i = 1}^{n} \pdeg_i$
. Furthermore, every element $\sigma \in \Sigma$ can be written in
a canonical form $\lambda + \sum_{i=1}^{t} w_i e_i,$ with the first
coordinate of $\lambda$ less than $\sum_{i=1}^{t}\pdeg_i$ and $w_1,
\ldots, w_t \in \bz$.
\end{prop}

@@

\begin{lem}
\label{lem:bound-ker-chi}
Retaining the notation of Lemma ~\ref{lem:composite-map}, if $\Sigma$ has extremal rays $e_1,\ldots, e_r$ in degrees $d_1, \ldots, d_r$ then $\ker \chi$ has a system of generators in degrees at most $2(\sum_{i=1}^{r}d_r-1)$.
\end{lem}
\begin{proof}
Since $e_1, \ldots, e_r$ are extremal rays, by Proposition ~\ref{prop:cone-generation}, every element $\sigma \in \Sigma$ can be written in a canonical form $\lambda + \sum_{i=1}^{r}e_i$. Now, letting $\lambda_0 := 0,\lambda_1, \ldots, \lambda_m$ be all elements of $\Sigma$ so that we can write $\lambda_i = \sum_{i=1}^{r}\bida_i e_i$ with $0 \leq \bida_i < 1$ for all $i$. Then, for any $1 \leq i \leq j \leq m,$ we can write $\lambda_i + \lambda_j$ in the above canonical form, yielding a relation in degree at most $\deg \lambda_i + \deg \lambda_j \leq 2 \cdot \left( \sum_{i=1}^{r}d_r -1 \right).$ Furthermore, these relations generate all relations, as one can apply a sequence of these relations to put any $\sigma \in \Sigma$ into canonical form.
\end{proof}

@@

\begin{rem}
\label{rem:ghost-motivation}
Example ~\ref{eg:hyperplane} shows that the naive generalization of ~\cite[Theorem 8]{dorney:canonical} of generation in degree at most $\sum_{i=1}^{n}\ell_i$ cannot possibly hold. The reason for this is that the divisors may ``be contained in a hyperplane,'' i.e., be expressable as functions in $m$ of the $m+1$ variables on $\bp^m$. This problem can easily be circumvented by adding in ``ghost points.'' That is, we may add divisors of the form $0 \cdot H_i$ to $D$, and reorder so that if $D = \sum_{i=1}^{n}\alpha_i V(f_i)$, then $f_0, \ldots, f_n$
\end{rem}

@@

We now move on offer bounds on generators and relations of $R_D$ for a general divisor $D$ on $\bp^m$.

Let $D = \sum_{i=1}^n \alpha_i C_i$ for curves $C_i$ on $\mathbb{P}^m$ and let $f_i$ be homogeneous polynomials in $x_0, \ldots, x_m$ such that $C_i = v(f_i)$ for all $i$; suppose $\deg(f_i) = a_i$. 
\begin{lem} \label{P-m-generators}
$R_D$ is generated in degree at most $\sum_{i=1}^n \ell_i a_i$.
% and relations in degree at most  .\todo{insert numbers}
\end{lem}
\begin{proof}
Let 
\begin{align}\label{eqn:Sigma-def}
	\Sigma = \{(d, c_1, \ldots, c_n) \in \bz^{n+1} : c_i \geq - d \alpha_i, \; 1 \leq i \leq n, \text{ and } \sum_{i=1}^{n}k_i^j = 0, 1 \leq j \leq t\}.
\end{align}
We note that $\Sigma$ has extremal rays given by the lattice points 
\begin{equation}\label{defn:e_i-P-n}
	e_i = (\ell_i \bida_i, - \alpha_1 \ell_i \bida_i, \ldots -\alpha_{i-1} \ell_i \bida_i, \ell_i \sum_{j\ne i} \alpha_j \bida_j, -\alpha_{i+1} \ell_i \bida_i, \ldots, -\alpha_n, \ell_i \bida_i)
\end{equation}
for each $i\in \{1, \ldots n\}$.
Viewed inside of $R_D$, we have $e_i$ corresponding to
\begin{equation}\label{eqn:epilon-def-P-m}
	\epsilon_i = \prod_{j\ne i} (f_j)^{-\alpha_j \ell_i a_i} {f_i}^{\ell_i \sum_{j\ne i} \alpha_j a_j}\in R_D.
\end{equation}
Therefore, applying proposition \ref{prop:cone-generation}, we see $R_D$ is generated in degrees less than
\[
	\sum_{i=1}^n \ell_i a_i.
\]
\end{proof}

Let $x_1, \ldots x_r$ be the generators in degrees at most $\sum_{i=1}^n \ell_i \bida_i$ (given by lemma \ref{P-m-generators}), and let $\phi$ be the surjection $k[x_1, \ldots x_r] \to R_D$.  We can factor $\phi$ through 
\[
	\Sigma =  \langle u^d z_1^{c_1} \cdots z_n^{c_n} : c_i \in \mathbb{Z}, \; c_i \geq -d \alpha_i, \mbox{ and }\sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{n} \bidb_i c_i. \rangle 
\]
by
\[
\begin{tikzcd}
\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
x_i \ar r & u^{d_i}z_1^{c_{i1}} \cdots z_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]

By Lemma \ref{lem:bound-ker-chi}, we can bound the degree of generation of $\ker(\chi)$ below
\[
	2 \sum_{i=1}^n \ell_i a_i.
\]
Finally, we calculate the degree of generation of $\psi$:

\begin{lem}\label{P-m-relations-psi}
$\ker(\psi)$ is generated in degree less than
\[
	\frac{\max_{1\le i \le n}(\bida_i)}{\deg(D)} +  \sum_{i=1}^n \ell_i a_i.
\]
\end{lem}
\begin{proof}
We first claim that $\psi$ there are $\beta_1, \ldots, \beta_n$ such that $\psi$ has relations of the form
\begin{equation}\label{eqn:relations-psi}
	(z_i - \beta_i)\prod_{j=1}^n {z_j}^{c_{j}}
\end{equation}

\noindent
for all $d\in \mathbb{N}$ and $c_i \ge -\alpha_i d$ satisfying $\bida_j + \sum_{i = 1}
^n \bida_i c_i = 0$.

To find such a $\beta$, we simply write $\epsilon_j$ as a 
polynomial in $\epsilon_1, \ldots \epsilon _3$. Furthermore, these
generate all relations, since they allow us to reduce any $\prod_{i =
1}^n \epsilon_i^{r_i}$ to a canonical form as given by Proposition \ref{prop:cone-generation}.

We now fix $z_i$ and bound the degree of generation of this set of relations of the form $(z_i - \beta_i) \prod_{j=1}^n z_j^{c_j}$.  We first associate $(z_i - \beta_i)\prod_{j=1}^n z_j^{c_j}$ with $(d, c_1, \ldots, c_n)$ where $d$ is its degree.  Let $\Sigma_i \subseteq \mathbb{Z}^{n+1}$ be the set of points of the form $(d, r_1, \ldots, r_n)$ satisfying $r_j \ge -d \alpha_j$ for all $j$ and $\sum_{j=1}^n r_j a_j = -a_i$.  
Then we see $\Sigma_i$ is contained in the $\mathbb{R}_{\ge 0}$-span of $e_1, \ldots, e_n$ (as defined in Equation \ref{defn:e_i-P-n}) translated by by $\delta_i = (\frac{a_i}{\deg(D)}, -\frac{\alpha_1 a_1}{\deg(D)}, \ldots, - \frac{\alpha_n a_n}{\deg(D)})$.

Therefore, we can write any element of $\Sigma'$ as
\[
	\delta_i + \sum_{j=1}^n r_j e_j
\]
where $r_j \in \mathbb{R}$ for each $j$.

Whenever some $r_j \ge 1$, we can generate the relation at hand by one of lower degree by factoring out a copy of $\epsilon_j$ (see Equation \ref{eqn:epilon-def-P-m}).  Therefore, $\ker(\psi)$ is generated in degrees less than
\[
	\frac{\bida_i}{\deg(D)} + \sum_{i=1}^n \ell_i a_i.
\]
\end{proof}

Therefore, since $\ker(\chi)$ and $\ker(\psi)$ are generated in degrees less than $2\sum_{i=1}^n \ell_i a_i$ and $\delta_i + \sum_{i=1}^n \ell_i a_i$ respectively, Lemma \ref{lem:composite-map} implies that $\ker(\phi)$ is generated in degrees less than
\[
	\max \left(2 \sum_{i=1}^n \ell_i a_i, \frac{\max_{1\le i \le n}(\bida_i)}{\deg(D)} + \sum_{i=1}^n \ell_i a_i \right).
\]

@@

\begin{thm}\label{P-m-generators/relations}
Let $D = \sum_{i=1}^n \alpha_i C_i$ for curves $C_i$ on $\mathbb{P}^m$.  Let $f_i$ be homogeneous polynomials in $x_0, \ldots, x_m$ such that $C_i = v(f_i)$ for all $i$; suppose $\deg(f_i) = a_i$. 
Then $R_D$ is generated in degree at most 
\[
	\sum_{i=1}^n \ell_i a_i
\]
with relations generated in degree at most
\[
	\max \left(2 \sum_{i=1}^n \ell_i a_i, \frac{\max_{1\le i \le n}(\bida_i)}{\deg(D)} + \sum_{i=1}^n \ell_i a_i \right).
\]
\end{thm}

\section{Canonical Rings on Hirzebruch surfaces}
Let $D=\sum_{i=1}^n \alpha_i C_i \in \mathbb{Q} \otimes \di(\hirz_
k)$ for some irreducible curves $C_1, \ldots C_n$ of $\hirz_k$, such
that $C_1, \ldots, C_4$ are independent curves with bi-degrees $(1,0)
, (1,0), (0,1)$, and $(0,1)$ respectively. Let $C_i = V(f_i)$ where
$f_i \in \mathscr{O}(\bida_i, \bidb_i)$.  Suppose $f_1$ and $f_2$ are independent linear
polynomials in $x_0$ and $x_1$ and $f_3$ and $f_4$ are independent
linear polynomials in $y_0$ and $y_1$.

\begin{defn}
Define 
\begin{equation}\label{eqn:define-T=(D)}
	\Te(D) = \left\{i \in \{1, \ldots, n\}: \bida_i \sum_{k=1}^n \bidb_k 
\alpha_k = \bidb_i \sum_{k=1}^n \bida_k \alpha_k \right\},
\end{equation}

\begin{equation}\label{eqn:define-T+(D)}
	\Tp(D) = \left\{ i \in \{1, \ldots, n\}:  \vphantom{\sum_{k=1}^n} 
\bida_i \sum_{k=1}^m \alpha_k \bidb_k > \bidb_i \sum_{k=1}^n \alpha_k \bida_k 
\right\},
\end{equation}

\noindent
and
\begin{equation}\label{eqn:define-T-(D)}
	\Tm(D) = \left\{ j \in \{1, \ldots, n\}: \bida_j \sum_{k=1}^n \alpha_
k \bidb_k < \bidb_j \sum_{k=1}^n \alpha_k \bida_k s \right\}.
\end{equation}
\end{defn}
